Critical Temperature of the Ising Ferromagnet on Proximity Graphs derived from Square Lattices by Site Displacement

### 2015-08-31 Tex/paper/abstract.tex
Paper Entwurf 2. Iteration (Christophs Anmerkungen)

We perform Monte Carlo simulations to determine the critical temperatures
of an Ising Ferromagnet (IFM) located on the sites of different types of
2D proximity graphs.
%~ In particular we consider Relative Neighborhood graphs and Gabriel graphs
%~ which are subgraphs of the Delaunay Triangulation for a given set of points.
The graphs are derived from square lattices where nodes are displaced by
a Gaussian distributed random variable.
%~ The limit of maximal disorder thus corresponds to a Poisson point process.
The deviation of the proximity graph from a square lattice is
governed by the width \(\sigma\) of the Gaussian distribution.
In our model, the coupling strength depends on the Euclidean distance
between the coupled spins.
The simulations are carried out on graphs with \(N=16^{2}\) to \(N=128^{2}\)
nodes utilizing the Wolff cluster algorithm and parallel tempering in a
wide temperature range around the critical point to measure the Binder Cumulant.
The critical temperatures are shown to depend mainly on the average degree
and the type of the underlying proximity graph.
We further verify that the model lies within the universality class of
the 2D IFM using finite-size scaling methods.

>>>>

We perform Monte Carlo simulations to determine the critical temperatures
of Ising Ferromagnets (IFM) on proximity graphs with different underlying
node-sets, which can be governed by a parameter $\sigma$.
The coupling strengths arise from the Euclidean distance
between coupled spins.
The simulations are carried out on graphs with \(N=16^{2}\) to \(N=128^{2}\)
nodes utilizing the Wolff cluster algorithm and parallel tempering in a
wide temperature range around the critical point to measure the Binder Cumulant.
The critical temperatures are shown to depend mainly on the average degree
and the type of the underlying proximity graph.
We further verify using finite-size scaling methods that the IFM on proximity
graphs is in the same universality class as the IFM on a two-dimensional
square lattice.

### 2015-08-31 Tex/paper/introduction.tex
Paper Entwurf 2. Iteration (Christophs Anmerkungen)

%~ Hallo Welt \cite{Ising1925}
%~ \blindtext
%~ ising wichtig, drosophila und so
%~ Janke und Lima haben das auch gemacht
%~ TODO: mehr finden

[todo: geo und bio cites] \cite{Karp2000} \cite{Santi2005} \cite{Bose2001} \cite{Kuhn2003} \cite{Selander1975} \cite{Sokal1978} \cite{Sokal1980}

The Ising model for a ferromagnet \cite{Ising1925} is
perhaps the most extensive studied model in statistical mechanics.
It consists of spins wich can point up or down and interact with
neighboring spins. It shows a second order phase transition between
a ferromagnetic, i.e.~all spins having the same value, and
paramagnetic phase. While there exist analytic solutions\cite{Onsager1944,Wannier1945}
describing the phase transition for Ising models on
regular lattices in two dimensions, higher dimensional lattices or
non-regular lattices are the domain of Monte Carlo simulations.

Some work \cite{Janke1994,Lima2000,Lima2008} was done to determine
the critcal temperatures and confirm the universality of 2D IFM
where the spins are arranged randomly in the unit square by a
Poisson point process and neighbor relationships are determined by a
Delaunay triangulation to maintain the planarity of the system.
This article extends the former research to other
irregular lattice types closely related to the Delaunay triangulation and
provides an interpolation between the square lattice IFM
and spins distributed randomly in the plane.
For a range of this interpolation the critcal temperatures are determined
and it is confirmed that configurations from this range belong to the
IFM universality class.

In section \ref{ssec:model} the method to distribute the sites of the
irregular lattice and in \ref{ssec:graphtypes} the rules to establish the
bonds between them is explained, section \ref{sec:results} will show the
details and results of our simulations, before section \ref{sec:conclusion}
closes this article with the conclusions and an outlook.

>>>>

The Ising model for a ferromagnet \cite{Ising1925} is perhaps the
most extensive studied model in statistical mechanics. It consists
of spins which can point up or down and interact with neighboring
spins. It shows a second order phase transition separating a
ferromagnetic and a paramagnetic phase. While there exist analytic
solutions\cite{Onsager1944,Wannier1945} describing the phase
transition for Ising models on regular lattices in two dimensions,
higher dimensional lattices and non-regular lattices are primarily studied
by Monte Carlo simulations.

Some work \cite{Janke1994,Lima2000,Lima2008} was done to determine
the critcal temperatures and confirm the universality of two-dimensional IFM
where the spins are arranged randomly in the unit square by a
Poisson point process and neighbor relationships are determined by a
Delaunay triangulation that maintains the planarity of the system.
This article extends the former research to other
irregular lattices, namely the Relative Neighborhood graph (RNG) and
Gabriel graph (GG), which belong, as the Delaunay triangulation, to the
family of \emph{proximity graphs}. To get some intuition for proximity
graphs, their purpose is to connect nodes which are close to each other.
Hence they are suited to generalize problems defined on regular lattices
with nearest neighbor relationships.
[todo: geo und bio cites] \cite{Karp2000} \cite{Santi2005} \cite{Bose2001} \cite{Kuhn2003} \cite{Selander1975} \cite{Sokal1978} \cite{Sokal1980}
This family of graphs have been applied in geographic variation studies in
biology and as potential candidate for ``virtual backbones'' in ad-hoc
networks. % also machine learning

Beginning with a square lattice the nodes can be shifted randomly governed
by a tunable parameter $\sigma$ until the distribution of the nodes is
indistinguishable from a Poisson point process.
For every realization of nodes, the construction rules for the proximity
graph at hand define an edge-set.
%~ The edge-set for any realization of node-placements is well defined by
%~ the construction rules of the
We measure the critical temperatures by varying $\sigma$ and confirm that
the universality class does not change in the process.

%~
 %~ closely related to the Delaunay triangulation and
%~ provides an interpolation between the square lattice IFM
%~ and spins distributed randomly in the plane.
%~ For a range of this interpolation the critcal temperatures are determined
%~ and it is confirmed that configurations from this range belong to the
%~ IFM universality class.

In Sec. \ref{ssec:model} the model and the underlying proximity graphs
are explained. Sec. \ref{sec:results} will show the
details and results of our simulations. Sec. \ref{sec:conclusion}
closes this article with the conclusion and an outlook.

### 2015-08-31 Tex/paper/model.tex
Paper Entwurf 2. Iteration (Christophs Anmerkungen)

\subsection{The Disordered Ising Model}
\label{ssec:model}
    We call the model under scrutiny in this article the \emph{disordered IFM}
    which is derived from the well understood \emph{square lattice IFM} \cite{Onsager1944}.
    That is a square lattice with lateral length \(L\) and \(N=L^2\)
    sites with periodic boundary conditions.
    Each site $i$ has an Ising spin \(s_i \in \{-1,+1\}\) and interacts with its
    nearest neighbors described by the Hamiltonian
    \begin{equation}
        \mathcal{H} = - \sum_{\avg{i,j}}J_{ij}s_{i}s_{j}.
        \label{eq:hamiltonian}
    \end{equation}
    \(\avg{i,j}\) refers to nodes \(i\) and \(j\) which are nearest
    neighbors, so that they are directly coupled to each other.
    This will be called an edge.
    The coupling between \(i\) and \(j\) is characterized by \(J_{ij} = J\),
    the coupling constant.

    The most important difference of the square lattice IFM to the
    disordered IFM is that the sites of the square lattice are
    displaced introducing geometric disorder resulting in a non regular graph
    structure.
    The displacement is randomly Gaussian distributed with standard
    deviation \(\sigma\), i.e.~the \(x\) and \(y\) coordinates of the
    sites are displaced by random \(\Delta x\) and \(\Delta y\) drawn
    from a Gaussian distribution.
    %~ \begin{equation}
        %~ f(\Delta x)=\frac{1}{\sqrt{2\pi}\sigma}\mathrm{e}^{-\frac{\Delta x^2}{2\sigma^2}}.
        %~ \label{eq:gauss}
    %~ \end{equation}
    %~ This is sketched in Fig.\ \ref{fig:displacement}.
    %~ \begin{figure}[htbp]
        %~ \centering
        %~ \includegraphics[width=0.45\textwidth]{images/Displacement}
        %~ \caption[Sketch how the Displacement works]
        %~ {
            %~ Sketch how the displacement of the nodes works. The nodes
            %~ get displaced by \(\Delta x\) and \(\Delta y\) drawn from the
            %~ distributions displayed next to the points. The original
            %~ square lattice is indicated by dashed lines.
        %~ }
        %~ \label{fig:displacement}
    %~ \end{figure}
    In the following \(\sigma\) will also be called \emph{disorder parameter}.
    If we take ``nearest neighbor" in the Euclidean meaning, most sites
    will only have one nearest neighbor after the
    displacement. If only the edges between these neighbors remained,
    the lattice would collapse to many very small clusters. But if one
    left the edges as they were before the displacement, the edges would
    cross -- at least for big displacements. The crossing of edges then
    would destroy the planar character of the model.
    To avoid this, a new edge set will be established after the displacement
    as shown in Fig.~\ref{fig:RNG_sigma} for example.
    \begin{figure}[htb]
        \centering
        \subfigure[$\sigma = 0.09$]
        {
            \label{sfig:RNG_sigma:0.09}
            \includegraphics[width=0.14\textwidth]{images/RNG/out009}
        }
        \subfigure[$\sigma = 0.15$]
        {
            \label{sfig:RNG_sigma:0.15}
            \includegraphics[width=0.14\textwidth]{images/RNG/out015}
        }
        \subfigure[$\sigma = 0.21$]
        {
            \label{sfig:RNG_sigma:0.21}
            \includegraphics[width=0.14\textwidth]{images/RNG/out021}
        }
        \caption[Examples of RNG for different $\sigma$]
        {
            Examples of proximity graphs (more precise, Relative Neighborhood
            graph see Sec. \ref{ssec:graphtypes}) for different $\sigma$.
            Connections which cross a periodic boundary are indicated
            by edges which connect a solid node to a dashed node.
        }
        \label{fig:RNG_sigma}
    \end{figure}
    The edges are constructed
    so that for a given configuration of points the resulting graph is an
    instance of a so called proximity graph, which will be explained in Sec.~\ref{ssec:graphtypes}.
    The coupling constant \(J_{ij}\) gets
    identified with edge weights. \(J_{ij}\) depends on the
    geometric distance between the connected pair of sites. More precise,
    the weight of an edge is
    \begin{equation}
        E_{ij} = J_{ij} = \mathrm{e}^{\alpha (1-d_{ij})}
        \label{eq:coupling}
    \end{equation}
    where \(d_{ij}\) is the Euclidean
    distance between the nodes \(i\) and \(j\). Following Ref.~\cite{Lima2000}
    the free parameter \(\alpha\) is set to \(\alpha = 0.5\).
    Note that for \(\sigma = 0\) all \(d_{ij} = 1\) and therefore all
    \(J_{ij} = 1\), so that the disordered IFM is reduced to the
    square lattice IFM.

\subsection{Proximity Graphs}
\label{ssec:graphtypes}
    A graph \(G(V,E)\) is a set of nodes \(V\) and edges \(E \subset V^{2}\).
    In the scope of this model, the nodes get identified with the sites of the
    lattice. If two nodes are connected by an edge, the sites are neighbors.\\
    The edges of \emph{proximity graphs} connect nodes which are by some metric near
    to each other. Hence they are suited to generalize problems defined on regular
    lattices with nearest neighbor relationships.
    To determine the distance we use the Euclidean
    metric on two-dimensional graphs, though in principle every metric in any
    dimension can be used.

    The \emph{Relative Neighborhood graph} (RNG) \cite{Toussaint1980} is
    a proximity graph. Two nodes \(i\) and \(j\) with distance $d_{ij}$
    will be connected, if no other node is in the \emph {lune}. The lune
    is defined as the intersection of two circles with radius \(r =
    d_{ij}\) and centers on \(i\) and \(j\). See also the hatched region
    in Fig.~\ref{fig:lunes}. An Example is given in Fig.~\ref{sfig:lunes:rng}.

@@

    The \emph{Gabriel graph} (GG) \cite{Gabriel1969} is a subgraph
    of the Delaunay triangulation \cite{Delaunay1934,Katajainen}
    (DT), i.e.~for the same set of nodes \(V\) the edge set of the
    DT is a superset of the edge set of the GG. Two nodes \(i\) and
    \(j\) with distance \(d_{ij}\) will be connected with an edge,
    if a circle with its center on half way between \(i\) and \(j\)
    and radius \(r = \frac{d_{ij}}{2}\) contains no other nodes. See
    also the cross hatched region from Fig.~\ref{sfig:lunes:def}. An
    Example is given in Fig.~\ref {sfig:lunes:gg}. Note that the
    lune of the GG is completely enclosed in the lune of the RNG,
    therefore the RNG is a subgraph of the GG.

@@

    % TODO: alles hierdrunter
    %~ \subsubsection{Construction}
        %~ The simplest way to construct these graphs is to test for each
        %~ pair of nodes if any other node lies in
        %~ the lune of the pair. That is of complexity \(O (N^3)\), because
        %~ there are \(N(N-1)\) pairs and for each \((N-2)\) nodes to test. So
        %~ the running time of a straight forward implementation would be of
        %~ order \(O(N^3)\).
        %~ To reduce the running time one can basically use this simple method but only test
        %~ the connectedness for nodes which are near to the lune and abort if
        %~ one node inside the lune is found. To determine which nodes are
        %~ near the lune one can subdivide the area in \(M\) \emph{cells} of size
        %~ \(L_c \times L_c\) and save for each cell a list with nodes lying
        %~ inside like presented in Ref.\ \cite{RNGCell} for the RNG and sketched in Fig.\ \ref{fig:cell}.
        %~ For this research \(M = N\) is used.\\
        %~ \begin{figure}[htbp]
            %~ \centering
            %~ \subfigure[not Connected Nodes][]{
                %~ \label{sfig:cell:big}
                %~ \includegraphics[width=0.2\textwidth]{images/CellBig}
            %~ }
            %~ \subfigure[Connected Nodes][]{
                %~ \label{sfig:cell:small}
                %~ \includegraphics[width=0.2\textwidth]{images/CellSmall}
            %~ }
            %~ \caption[Sketch how the Cell Method works]
            %~ {
                %~ Sketch how the cell algorithm for the construction of the
                %~ RNG and GG works. Here with the lune definition of the GG.
                %~ The bounding box of the lune, which determines which cells have
                %~ to be tested, is marked with thick colored lines.
                %~ Left: shows that it is sufficient to find a
                %~ single node in an inner cell of the bounding box to discard
                %~ the edge between two distant nodes. The nodes inside the
                %~ other 8 marked cells do not have to be tested anymore.
                %~ Right: shows that only the nodes inside the
                %~ marked cell have to be tested, because there are no other nodes,
                %~ the edge can be drawn.
            %~ }
            %~ \label{fig:cell}
        %~ \end{figure}\\
        %~ Given the discretized cell structure it is just necessary to test the nodes in the cells which
        %~ resemble a rectangular bounding box of the lune. Most pairs will be
        %~ far away from each other and there will be one or more cells in the
        %~ middle of the bounding box, which are completely inside the lune.
        %~ Then only one node has to be tested from such a cell to discard an
        %~ edge between the nodes. On the other hand, nodes that will be connected
        %~ are near to each other so that only very few cells intersect the bounding
        %~ box of their lune and consequently only very few nodes have to be tested.\\
        %~ The complexity of this method should be between \(O(N^2)\) in the
        %~ best case, because for every pair at least one check has to be performed,
        %~ and \(O(N^3)\) in the worst case, e.g.\ all nodes are inside one
        %~ cell.\\
        %~ Note that even faster algorithms exsist for two dimensional RNG and GG, which require to first
        %~ generate a DT and use the fact, that the RNG and GG are subgraphs.
        %~ (For the construction of the DT for a given set of points one might
        %~ use existing software libraries, e.g.\ the \texttt{qHull}\cite{qhull} library.)\\
        %~ However, generation of the graphs is not time critical, because
        %~ the MC simulations take much longer. So only the simple cell algorithm
        %~ was used to gegnerate the graphs.\\

>>>>

\subsection{The Ising Model}
\label{ssec:model}
% \cite{Onsager1944}
    Commonly, the IFM is studies on a square lattice with lateral length
    \(L\) and \(N=L^2\) nodes with periodic boundary conditions.
    Each site $i$ has an Ising spin \(s_i \in \{-1,+1\}\) and interacts with its
    nearest neighbors described by the Hamiltonian
    \begin{equation}
        \mathcal{H} = - \sum_{\avg{i,j}}J_{ij}s_{i}s_{j}.
        \label{eq:hamiltonian}
    \end{equation}
    \(\avg{i,j}\) refers to nodes \(i\) and \(j\) which are
    nearest neighbors, i.e.~they are connected by an edge.
    The coupling strength between \(i\) and \(j\) is given by
    \(J_{ij}\).

    In this article the nodes of the square lattice are
    displaced introducing geometric disorder resulting in a non regular graph
    structure.
    The displacement is randomly Gaussian distributed with standard
    deviation \(\sigma\), i.e.~the \(x\) and \(y\) coordinates of the
    sites are displaced by random \(\Delta x\) and \(\Delta y\) drawn
    from a Gaussian distribution with zero mean.
    In the following \(\sigma\) will be called \emph{disorder parameter}.
    If we took ``nearest neighbor" in the Euclidean meaning, most sites
    would only have one nearest neighbor after the
    displacement. If only the edges between these neighbors remained,
    the lattice would collapse to many very small clusters. If on the
    other hand all edges remained as they were before the displacement,
    edges might cross each other -- at least for big displacements.
    The crossing of edges would destroy the planar character of the graph.
    To avoid this, a new edge set will be established after the displacement
    as shown in Fig.~\ref{fig:RNG_sigma}.
    \begin{figure}[htb]
        \centering
        \subfigure[$\sigma = 0.09$]
        {
            \label{sfig:RNG_sigma:0.09}
            \includegraphics[width=0.14\textwidth]{images/RNG/out009}
        }
        \subfigure[$\sigma = 0.15$]
        {
            \label{sfig:RNG_sigma:0.15}
            \includegraphics[width=0.14\textwidth]{images/RNG/out015}
        }
        \subfigure[$\sigma = 0.21$]
        {
            \label{sfig:RNG_sigma:0.21}
            \includegraphics[width=0.14\textwidth]{images/RNG/out021}
        }
        \caption[Examples of RNG for different $\sigma$]
        {
            Examples of proximity graphs (more precise, Relative Neighborhood
            graph see Sec. \ref{ssec:graphtypes}) for different $\sigma$.
            Connections which cross a periodic boundary are indicated
            by edges which connect a solid node to a dashed node.
        }
        \label{fig:RNG_sigma}
    \end{figure}
    For a given configuration of points the resulting graph is an
    instance of a so called proximity graph, which will be explained in Sec.~\ref{ssec:graphtypes}.
    The coupling constants \(J_{ij}\), i.e.~edge weights, depend on the
    geometric distance between the connected pair of sites. More precise,
    the weight of an edge is
    \begin{equation}
        E_{ij} = J_{ij} = \mathrm{e}^{\alpha (1-d_{ij})},
        \label{eq:coupling}
    \end{equation}
    where \(d_{ij}\) is the Euclidean
    distance between node \(i\) and \(j\). Following Ref.~\cite{Lima2000}
    the free parameter \(\alpha\) is set to \(\alpha = 0.5\).
    Note that for \(\sigma = 0\) the resulting \(d_{ij} = 1\) and therefore
    \(J_{ij} = 1\) for every edge $\{i,j\}$. Consequently, in this case
    the graph is a common square lattice.

\subsection{Proximity Graphs}
\label{ssec:graphtypes}
    A graph \(G(V,E)\) consists of a set of nodes \(V\) and edges \(E \subset V^{2}\).
    Two nodes connected by an edge are neighbors.
    Proximity graphs are defined in a metric space. In this article a
    two-dimensional space with the Euclidean metric is chosen, because
    it is the most common case and easy to visualize, though in principle
    every metric in any dimension can be used.

    The \emph{Relative Neighborhood graph} (RNG) \cite{Toussaint1980} is
    a proximity graph. Two nodes \(i\) and \(j\) with distance $d_{ij}$
    will be connected, if no other node is in the \emph {lune}. The lune
    is defined as the intersection of two circles with radius \(r =
    d_{ij}\) and centers on \(i\) and \(j\). That means that the edge
    $\{ij\}$ is part of the graph, if the condition
    \[d_{ij} \le \max\left{ d_{ik} + d_{kj} \right} \forall k \in V\setminus\{i,j\}\]
    is fulfilled. See also the hatched region
    in Fig.~\ref{fig:lunes}. An example is given in Fig.~\ref{sfig:lunes:rng}.

@@

    The \emph{Gabriel graph} (GG) \cite{Gabriel1969} is a subgraph
    of the Delaunay triangulation \cite{Delaunay1934,Katajainen}
    (DT), i.e.~for the same set of nodes \(V\) the edge set of the
    DT is a superset of the edge set of the GG. Two nodes \(i\) and
    \(j\) with distance \(d_{ij}\) will be connected with an edge,
    if a circle with its center on half way between \(i\) and \(j\)
    and radius \(r = \frac{d_{ij}}{2}\) contains no other nodes, i.e.~the
    condition
    \[d_{ij} < d_{ik} + d_{kj} \forall k \in V\setminus\{i,j\}\]
    is fulfilled. See
    also the cross hatched region from Fig.~\ref{sfig:lunes:def}. An
    example is given in Fig.~\ref {sfig:lunes:gg}. Note that the
    lune of the GG is completely enclosed in the lune of the RNG,
    therefore the RNG is a subgraph of the GG.

@@
